### 2021-11-16 documents/sections/verbs.tex
'to say'

\ex<kati> \apalai\\
\obj{kaʃi-ko} \qu{say!}\\
\parencite[35]{koehn1986apalai}
\xe
%
A comparison of the longest forms attested in each language is shown in \cref{tab:say}.
Segments of cognate forms in this and other tables were aligned automatically with LingPy \parencite{lingpy268}, for easier recognition of correspondences.
%
\input{floats/say.tex}
%
This verb was not affected by any of the extensions found in \PPek, \PWai, \PTir, \akuriyo, or \carijo (\crefrange{sec:pekodian}{sec:carijo}).
We do not know the first person prefix of its \yukpa reflex \obj{ka}.

As briefly mentioned in \cref{sec:waiwaian}, \textcite{hixkaryanaderby1985} analyzes this verb as transitive in \hixka.
This analytical choice is not only motivated by avoiding an idiosyncratic intransitive first person prefix \obj{ɨ-}:
\hixka \obj{ka} also behaves like a transitive verb in other ways, for instance by showing the complementary distribution of the third person marker \obj{n-} and preceding objects -- in this case direct speech or ideophones \exref{hixsay}.

\pex<hixsay>\hixka
\a<hix-119>
\begingl
\glpreamble onɨ wyaro nkekonɨ bɨryekomo, tɨyonɨ wya//
\gla onɨ wjaro n-ka-jakonɨ bɨrʲekomo tɨ-jonɨ wja//
\glb this like \gl{3}-say-\gl{rem}.\gl{cont} boy \gl{cor}-mother \gl{obl}//
\glft \qu{This is what the boy said to his mother.} \parencite[][36]{hixkaryanaderby1985}//
\endgl
\a<ordpl-35>
\begingl
\glpreamble moro ha, ketxkoná hatá.//
\gla moro ha ka-jat͡ʃkonɨ hatɨ//
\glb \gl{med}.\gl{dem}.\gl{inan} \gl{ints} say-\gl{rem}.\gl{cont}.\gl{pl} \gl{hsy}//
\glft \qu{“That one there” they said.} \parencite[][14]{derbyshire1965textos}//
\endgl
\xe
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}).
In \exref{hixsay.ordpl-35}, it does not occur, because \qu{they said} is preceded by direct speech.
This complementary distribution is otherwise only found with transitive verbs \parencite[59--60]{gildea1998}.
The verb shows the same pattern, albeit inconsistently, in \trio \parencite[267]{triocarlin2004}.

Further comparative evidence also points to \rc{ka[ti]} \qu{to say} showing transitive traits:
\trio \obj{ka} is characterized as the only intransitive verb being able to take the causative suffix \obj{-po} and the agentive nominalizer \obj{-ne} \parencite[263, 169]{triomeira1999}.
The exceptionality of \obj{ka} \qu{to say} taking \obj{-po} \qu{\gl{caus}.\gl{tr}} has also been noted for \kalina \parencite[82]{courtz2008carib} and \wayana \parencite[258]{wayanatavares2005}.
The agent nominalizer \rc{-ne} gave rise to the \panare gnomic verbal suffix \obj{-ɲe} on transitive verbs \parencite[184--185]{gildea1998}.
The occurrence of \obj{-ɲe} on \obj{ka} has led to differing categorizations as transitive \parencite[214]{panarepayne2013} or intransitive \parencite[102]{mattei1994diccionario}.
Reflexes of another transitive causativizer \rc{-metipo} \parencite{gildea2015valency} are found with \obj{ka} in \apalai \parencite[51]{koehn1986apalai} and \waiwai \parencite[52]{waiwaihawkins1998}.

Our classification of \qu{to say} as an intransitive verb is supported primarily by prefix patterns:
\kalina offers a minimal pair between transitive \obj{ka} \qu{to remove} and intransitive \obj{ka} \qu{to say}, \obj{sikai} \qu{I took it away} vs \obj{wɨkai} \qu{I said} \parencite[288, 45]{courtz2008carib}.\footnote{Interestingly, the \kalina causativized form \obj{kapo} \qu{to make say} does not have the regular \gl{1}>\gl{3} prefix \obj{s(i)-}, but irregular \obj{w(ɨ)-} \parencite[430]{courtz2008carib}.}
Similarly, \qu{to say} in Pekodian languages has a reflex of \gl{1}\gl{s} \rc{w-} \pcref{sec:pekodian}, and not \gl{1}>\gl{3} \obj{s-} (\bakairi) or \rc{ini-} (\PXin).
Additionally, the \gl{s_a_} class marker \obj{w-} occurs on nominalizations in \kalina \exref{kar-84}, and it is probably reflected in vowel length in the \trio \parencite[333]{triomeira1999} and \wayana \parencite[196]{wayanatavares2005} participles.

\ex<kar-84>\kalina \parencite[][202]{courtz2008carib}\\
\begingl
\glpreamble Òmakon \`wa oti ywykàpo kaiko.//
\gla o-ʔma-kon ʔwa oti ɨ-wɨ-ka-ʔpo kai-ko//
\glb \gl{2}-child-\gl{pl} \gl{obl} greeting \gl{1}-\gl{s_a_}-say-\gl{pst}.\gl{nmlz} say-\gl{imp}//
\glft \qu{Pass my greetings to your children.}//
\endgl
\xe
%
Summing up, this verb could be reconstructed as being intransitive based on its prefixes, but transitive based on some suffixes.
\hixka has lost the main intransitive criteria, making its reflex look more like a transitive verb.
It is not attested as being affected by any of the person marker extensions under discussion.

>>>>

\ex<kati> \apalai\\
\obj{kaʃi-ko} \qu{say!}\\
\parencite[35]{koehn1986apalai}
\xe
%
\cref{tab:say} shows a comparison of the longest attested forms for each language.
Cognate segments in \cref{tab:say,tab:bathe} were aligned automatically with LingPy \parencite{lingpy268}, for easier recognition of correspondences.
%
\input{floats/say.tex}%
%
This verb was not affected by any of the extensions found in \PPek, \PWai, \PTir, \akuriyo, or \carijo (\crefrange{sec:pekodian}{sec:carijo}).
We do not know the first person form of its \yukpa reflex \obj{ka}.

As briefly mentioned in \cref{sec:waiwaian}, \textcite{hixkaryanaderby1985} analyzes this verb as transitive in \hixka.
This analytical choice is not only motivated by the desire to avoid an idiosyncratic intransitive first person prefix \obj{ɨ-} instead of \obj{kɨ-}.
\hixka \obj{ka} also shows the complementary distribution of third person  \obj{n-} and preceding objects typical of transitive verbs in Cariban \parencite[60--81]{gildea1998}.
Due to its semantics, these objects are either direct speech \exref{hixsay} or ideophones.

\pex<hixsay>\hixka
\a<hix-119>
\begingl
\glpreamble onɨ wyaro nkekonɨ bɨryekomo, tɨyonɨ wya//
\gla onɨ wjaro n-ka-jakonɨ bɨrʲekomo tɨ-jonɨ wja//
\glb this like \gl{3}-say-\gl{rem}.\gl{cont} boy \gl{cor}-mother \gl{obl}//
\glft \qu{This is what the boy said to his mother.} \parencite[][36]{hixkaryanaderby1985}//
\endgl
\a<ordpl-35>
\begingl
\glpreamble moro ha, ketxkoná hatá.//
\gla moro ha ka-jat͡ʃkonɨ hatɨ//
\glb \gl{med}.\gl{dem}.\gl{inan} \gl{ints} say-\gl{rem}.\gl{cont}.\gl{pl} \gl{hsy}//
\glft \qu{“That one there” they said.} \parencite[][14]{derbyshire1965textos}//
\endgl
\xe
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}).
In \exref{hixsay.ordpl-35}, it does not occur, because \qu{they said} is preceded by direct speech.
At least the \trio cognate shows the same pattern, albeit inconsistently \parencite[267]{triocarlin2004}.

Comparative evidence also points to \rc{ka[ti]} \qu{to say} showing transitive traits:
\trio \obj{ka} is characterized as the only intransitive verb being able to take the causative suffix \obj{-po} and the agentive nominalizer \obj{-ne} \parencite[263, 169]{triomeira1999}.
The exceptionality of \obj{ka} \qu{to say} taking \obj{-po} \qu{\gl{caus}.\gl{tr}} has also been noted for \kalina \parencite[82]{courtz2008carib} and \wayana \parencite[258]{wayanatavares2005}.
The agent nominalizer \rc{-ne} gave rise to the \panare gnomic verbal suffix \obj{-ɲe} on transitive verbs \parencite[184--185]{gildea1998}.
The occurrence of \obj{-ɲe} on \obj{ka} likely led \textcite[214]{panarepayne2013} to categorize it as transitive, contrastive with the intransitive analysis by \textcite[102]{mattei1994diccionario}.
Finally, reflexes of the causativizer \rc{-metipo}, usually restricted to transitive verbs \parencite{gildea2015valency}, are found with \obj{ka} in \apalai \parencite[51]{koehn1986apalai} and \waiwai \parencite[52]{waiwaihawkins1998}.

Our categorization of \qu{to say} as an intransitive verb is supported primarily by its person prefixes.
\kalina offers a minimal pair between transitive \obj{ka} \qu{to remove} and intransitive \obj{ka} \qu{to say}, \obj{sikai} \qu{I took it away} vs \obj{wɨkai} \qu{I said} \parencite[288, 45]{courtz2008carib}.\footnote{Interestingly, the \kalina causativized form \obj{kapo} \qu{to make say} does not have the regular \gl{1}>\gl{3} prefix \obj{s(i)-}, but irregular \obj{w(ɨ)-} \parencite[430]{courtz2008carib}.}
Similarly, \qu{to say} in Pekodian languages has a reflex of \gl{1}\gl{s} \rc{w-} \pcref{sec:pekodian}, and not \gl{1}>\gl{3} \obj{s-} (\bakairi) or \rc{ini-} (\PXin).
Additionally, languages which differentiate direct prefixes from \gl{s_a_} prefixes by the presence of \obj{i} \parencite[495]{meira2010origin} show no \obj{i} for this verb, see \exref{sayintr.aku-154} and  \textcites[294]{triomeira1999}[195]{wayanatavares2005}[288]{ikpengpacheco2001}[150]{alves2017arara}[168]{hoff1968carib}.
Finally, the \gl{s_a_} class marker \obj{w-} occurs on nominalizations in \kalina \exref{sayintr.kar-84}, and it is probably reflected as vowel length in the \trio \parencite[333]{triomeira1999} and \wayana \parencite[196]{wayanatavares2005} participles.

\pex<sayintr>
\a<aku-154> \akuriyo \parencite[][113]{meira1998proto}\\
\begingl
\gla mɨ-ka//
\glb \gl{2}-say//
\glft \qu{You said.}//
\endgl
\a<kar-84> \kalina \parencite[][202]{courtz2008carib}\\
\begingl
\glpreamble Òmakon `wa oti ywykàpo kaiko.//
\gla o-ʔma-kon ʔwa oti ɨ-wɨ-ka-ʔpo kai-ko//
\glb \gl{2}-child-\gl{pl} \gl{obl} greeting \gl{1}-\gl{s_a_}-say-\gl{pst}.\gl{nmlz} say-\gl{imp}//
\glft \qu{Pass my greetings to your children.}//
\endgl
\xe

Summing up, this verb could be reconstructed as being intransitive based on its prefixes, but transitive based on some suffixes.
\hixka has lost the main intransitive criteria, making its reflex look more like a transitive verb.
It is not attested as being affected by any of the person marker extensions  in \cref{sec:extensions}.
